---
title: "Pre-processing pupil data"
author: "Adaptive automation team"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(eyelinker)
library(PupilPre)
library(plyr)
library(dplyr)
source("./src/preprocessing_helper_functions.R")
```

## Pre-processing pupil data

For pre-processing we rely on the pupilpre package and follow the steps outlined
in the vignette for this package (see links below). Since we have .asc files
and pupilpre expects a sample report some functions are sourced that bring
the .asc files in a format that pupilpre can work with.

```{r}
# Load in data from .asc file
# Steps taken through out the entire file follow mostly the Basic preprocessing of pupil size data vignette
# see: https://cran.microsoft.com/snapshot/2020-07-25/web/packages/PupilPre/vignettes/PupilPre_Basic_Preprocessing.html

path_to_sub <- "SPECIFY_PATH_HERE"
label <- "PROVIDE_DESCRIPTIVE_LABEL_HERE" # e.g. UM1_automated
recorded_eye <- "Left" # do not change this :)

sample_cl <- asc_to_ppl_pre(path_to_sub,
                               label,
                               recorded_eye)



# Align to experimental begin as signaled by AUTOMATION_DECREASE none,EXPERIMENT STARTED msg
sample_cl <- align_msg(sample_cl,Msg="^AUTOMATION_DECREASE none,EXPERIMENT STARTED")

# Create time series object (we don't need to adjust so this can just be copied)
sample_cl$Time <- sample_cl$Align

# Create eye-agnostic data columns
sample_cl <- ppl_select_recorded_eye(sample_cl,Recording = "L")

# Marking & removing data-points outside screen
sample_cl <- recode_off_screen(data=sample_cl,ScreenSize = c(1920,1080))
sample_cl$Pupil[!(sample_cl$Screen == "OnScreen")] <- NA
  
# Plot before blink removal
plot(sample_cl$Time,sample_cl$Pupil,ylab = "Pupil",xlab = "Time",type="l")

# de-blink via SR logs
deblinked_sample <- clean_blink(sample_cl,Delta = 5,LogFile = "./BlinkCleanup.rds")

# Verify de-blinking (this takes some time to load the data)
verify_cleanup_app(deblinked_sample, LogFile = "./BlinkCleanup.rds")

# Final plot
plot(deblinked_sample$Time,deblinked_sample$Pupil,ylab = "Pupil",xlab = "Time",type="l")
```

### Manual pre-processing

We removed left-over blinks using the user clean app as described in the
de-blinking vignette (see link below)

```{r}
# Manual Clean up preparations.
# We want to follow the steps outlined in the vignette below:

# See: https://cran.microsoft.com/snapshot/2020-07-25/web/packages/PupilPre/vignettes/PupilPre_Cleanup.html
# However, since we have 10 minutes of re-cordings we first split this up
# into 5 second windows.

data_for_manual_clean <- deblinked_sample

windows <- rep(1:200,each=2500)
data_for_manual_clean$windows <- windows[1:nrow(data_for_manual_clean)]
par(mfrow=c(2,2))
for (window in 1:max(data_for_manual_clean$windows)) {
  plot(data_for_manual_clean$Time[data_for_manual_clean$windows == window],
       data_for_manual_clean$Pupil[data_for_manual_clean$windows == window],
       ylab="Pupil",
       xlab="Time",
       type="l",
       main=window,
       ylim=c(min(data_for_manual_clean$Pupil,na.rm=T),
              max(data_for_manual_clean$Pupil,na.rm=T)))
}

# re-write event column so that we don't have to try and clean a
# 10 minutes stream of pupil data sampled at 500HZ...
oldEvent <- data_for_manual_clean$Event
data_for_manual_clean$Event <- as.factor(data_for_manual_clean$windows)
```

```{r}
# Now we perform manual clean up based on clear deviations identified in
# the plots above.
user_cleanup_app(data_for_manual_clean,
                 LogFile = paste0("./",label,"_ManualCleanup.rds"))
```

```{r}
# Apply clean-up (or if you want to algin your data differently and have
# completed pre-processing before just set the if(T) to if(F) and load
# in the final cleaned data)
if(T){
  final_sample <- safe_apply_manual_clean(data_for_manual_clean,
                                          paste0("./",label,"_ManualCleanup.rds"))
  # Rewrite event column split earlier
  final_sample$Event <- oldEvent

  saveRDS(final_sample,file=paste0("./",label,"_final_cleaned.RDS"))
} else{
  final_sample <- readRDS(file=paste0("./",label,"_final_cleaned.RDS"))
}

# Final plot
plot(final_sample$Time,final_sample$Pupil,ylab = "Pupil",xlab = "Time",type="l")
```

# Aligning data for analysis

The next steps were dependent on what we wanted to investigate:

- changes in pupil size over the entire session
- changes in pupil size following a question
- for automated blocks:
  - changes in pupil size following a change in automation
  
To achieve this we again just follow the remaining steps in the aforementioned
vignette: baselining and down-sampling. However, depending on the question we
first re-align the data.

First: changes in pupil size over the entire session

```{r}
# Baselined data
baselined_sample <- baseline(final_sample,
                             BaselineWindow = c(-200,200),
                             BaselineType = "Subtraction")

plot(baselined_sample$Time,baselined_sample$Pupil,
     ylab = "Pupil change relative to baseline",
     xlab = "Time",
     type="l")
```

```{r}
# Downsample
downsampled_data <- downsample(baselined_sample,
                               SamplingRate = 500,
                               NewRate = 50)
# save data
saveRDS(baselined_sample,file=paste0("./",label,"_baselined_dat.RDS"))
saveRDS(downsampled_data,file=paste0("./",label,"_downsampled_dat.RDS"))

plot(downsampled_data$Time,downsampled_data$Pupil,
     ylab = "Pupil change relative to baseline and at 50HZ",
     xlab = "Time",
     type="l")

# We can still plot message lines, here for questions.
for(row_index in 1:nrow(baselined_sample)){
  if(!is.na(baselined_sample$SAMPLE_MESSAGE[row_index])){
    msg_txt <- baselined_sample$SAMPLE_MESSAGE[row_index]
    msg_time <- baselined_sample$Time[row_index]
    if(startsWith(as.character(msg_txt),"QUESTION") ||
       startsWith(as.character(msg_txt),"NEXT BLOCK,QUESTION")) {
      abline(v=msg_time,col="green",lty=2)
    }
  }
}

```

Second: changes in pupil following a question

```{r}
# Create new data-frame
questionDatCleaned <- final_sample

# Reset message to only Question asked for regular expression to work.
questionDatCleaned$SAMPLE_MESSAGE <- as.character(questionDatCleaned$SAMPLE_MESSAGE)
questionDatCleaned$SAMPLE_MESSAGE[startsWith(questionDatCleaned$SAMPLE_MESSAGE,"NEXT BLOCK,QUESTION ASKED 1")] <- "QUESTION ASKED 1"
questionDatCleaned$SAMPLE_MESSAGE[startsWith(questionDatCleaned$SAMPLE_MESSAGE,"ERROR MESSAGES LOST 1,QUESTION ASKED 1")] <- "QUESTION ASKED 1"
questionDatCleaned$SAMPLE_MESSAGE <- as.factor(questionDatCleaned$SAMPLE_MESSAGE)

# Also mark for each question whether an AUTOMATION_INCREASE was received before
questionDatCleaned <- extract_samples_before_after_msg(questionDatCleaned,
                                                       "QUESTION ASKED",
                                                       0,
                                                       7500,
                                                       split_msg_by_block=T,
                                                       split_msg_decision=10,
                                                       check_pre_msg=T,
                                                       check_pre_msg_val="AUTOMATION_INCREASE",
                                                       check_pre_msg_inval="AUTOMATION_DECREASE"
                                                       )

# Re-align
questionDatCleaned <- align_msg(questionDatCleaned,Msg="^QUESTION ASKED")
questionDatCleaned$Time <- questionDatCleaned$Align

```

```{r}
# baseline again
question_baselined_sample <- baseline(questionDatCleaned,
                             BaselineWindow = c(0,200),
                             BaselineType = "Subtraction")

unique(question_baselined_sample$Baseline)

plot(question_baselined_sample$Time[question_baselined_sample$Event == "1.1"],
     question_baselined_sample$Pupil[question_baselined_sample$Event == "1.1"],
     type="l",ylab = "Pupil baselined",xlab="Time following question (in ms)")
abline(v=0,lty=2,col="red")
```

```{r}
# Downsample again
question_downsampled_data <- downsample(question_baselined_sample,
                               SamplingRate = 500,
                               NewRate = 50)
# save data
saveRDS(question_baselined_sample,file=paste0("./",label,"_question_baselined_dat.RDS"))
saveRDS(question_downsampled_data,file=paste0("./",label,"_question_downsampled_dat.RDS"))

avgDat <- ddply(question_downsampled_data,c("Time"),summarise,
                meanPupil = mean(Pupil,na.rm=T),
                se=(sd(Pupil,na.rm = T)/length(Pupil)))

plot(avgDat$Time,avgDat$meanPupil,type = "l",lwd=2)
lines(avgDat$Time,avgDat$meanPupil + (1.96*avgDat$se),lty=2)
lines(avgDat$Time,avgDat$meanPupil - (1.96*avgDat$se),lty=2)
```

Finally: changes in pupil following a an automation change.
Important: This only works for the automated session data, since the AAS
did not write any messages during the un-aided session!

Here we can be creative, for now I only collect 1 to 2. If you want to
collect for decrease or increase just adjust the AUTOMATION_INCREASE messages
in line 243 and 249 accordingly.

```{r}
# Create new data-frame
changeDatCleaned <- final_sample

changeDatCleaned <- extract_samples_before_after_msg(changeDatCleaned,
                                                       "AUTOMATION_INCREASE",
                                                       0,
                                                       5000
                                                       )


changeDatCleaned <- align_msg(changeDatCleaned,Msg="^AUTOMATION_INCREASE")
changeDatCleaned$Time <- changeDatCleaned$Align
```

```{r}
# baseline again
change_baselined_sample <- baseline(changeDatCleaned,
                             BaselineWindow = c(0,200),
                             BaselineType = "Subtraction")

unique(change_baselined_sample$Baseline)

plot(change_baselined_sample$Time[change_baselined_sample$Event == "1"],
     change_baselined_sample$Pupil[change_baselined_sample$Event == "1"],
     type="l",ylab = "Pupil baselined",xlab="Time following change (in ms)")
abline(v=0,lty=2,col="red")
```

```{r}
# Downsample again
change_downsampled_data <- downsample(change_baselined_sample,
                               SamplingRate = 500,
                               NewRate = 50)
# save data
saveRDS(change_baselined_sample,file=paste0("./",label,"_change_baselined_dat.RDS"))
saveRDS(change_downsampled_data,file=paste0("./",label,"_change_downsampled_dat.RDS"))

avgDat <- ddply(change_downsampled_data,c("Time"),summarise,
                meanPupil = mean(Pupil,na.rm=T),
                se=(sd(Pupil,na.rm = T)/length(Pupil)))

plot(avgDat$Time,avgDat$meanPupil,type = "l",lwd=2,
     ylab="Average change in pupil size",
     xlab = "Time after automation increase (in ms)")
lines(avgDat$Time,avgDat$meanPupil + (1.96*avgDat$se),lty=2)
lines(avgDat$Time,avgDat$meanPupil - (1.96*avgDat$se),lty=2)
abline(v=0,lty=2,col="red")
```
