---
title: "AMMS of pupil size"
author: "Adaptive automation team"
date: "10/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mgcv) # 1.8-36
library(itsadug) # 2.4
library(plyr)
library(dplyr)
```

This file fits the Additive mixed effect models (AMMs, see Wood, 2017)
for the three data-sets we created earlier:

- changes in pupil size over the entire session
- changes in pupil size following a question
- for automated blocks:
  - changes in pupil size following a change in automation
  
### Loading and merging data from all subjects

First for entire sessions

```{r}
# Specify path and pattern
path_to_folders <- "C:/Users/Chiel/Documents/CCS/User Models/cleaned_dat/"
data_patterns <- c("_automated_downsampled_dat.RDS","_non_automated_downsampled_dat.RDS")
condition_names <- c("aided","unaided")

# collect subjects
subjects <- list.files(path_to_folders)[startsWith(list.files(path_to_folders),"UM")]

align_start <- -500
align_end <- 600000

# Check alignments
for(subj in subjects) {
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    if(min(sub_dat$Time,na.rm=T) > align_start) {
      align_start <- min(sub_dat$Time,na.rm=T)
    }
    if(max(sub_dat$Time,na.rm=T) < align_end) {
      print(max(sub_dat$Time,na.rm=T))
      print(subj)
      align_end <- max(sub_dat$Time,na.rm=T)
    }
    rm(sub_dat)
  }
  
}


# Merge data
merged_dat <- NULL
for(subj in subjects) {
  condition_index <- 1
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    cat(subj,data_pattern,"\n")

    sub_dat <- sub_dat[((sub_dat$Time >= align_start) & (sub_dat$Time <= align_end)),]
    
    # Assign condition
    sub_dat$condition <- condition_names[condition_index]
    condition_index <- condition_index + 1
    sub_dat$subject_c <- subj
    
    if(length(colnames(sub_dat)) != length(colnames(merged_dat)) && subj != "UM1") {
      cat(subj, data_pattern, " had too many columns.\n")
      # This happens in case the wrapper function (safe_apply_manual_clean) had to intervene
      sub_dat <- sub_dat[,!(colnames(sub_dat) %in% c("PupilClean","PupilPrevious"))]
    }
    
    merged_dat <- rbind(merged_dat,sub_dat)
    rm(sub_dat)
  }
}

# set to factor for gams
# Note: the Subject column is actually a subject*session interaction
# subject_c is truly a subject column
merged_dat$condition <- as.factor(merged_dat$condition)
merged_dat$subject_c <- as.factor(merged_dat$subject_c)

# Averages plots
avgDat <- ddply(merged_dat,c("Time","condition"),summarise,
                meanPupil = mean(Pupil,na.rm=T),
                se=(sd(Pupil,na.rm = T)/length(Pupil)))


plot(0,0,type="n",ylim=c(-500,300),xlim=c(0,600000),
     ylab="Average change in pupil size",
     xlab = "Time (in ms)")

colrs <- c("red","blue")
col_index <- 1
for(condition in condition_names) {
  print(condition)
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition],lwd=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] + (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] - (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  col_index <- col_index + 1
}
abline(v=0,lty=2,col="red")

```

### Fitting models

```{r}
# build base model based on the predictors outlined in (Van Rij, 2019)
# to compare conditions.
m1.1 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=10) +
            te(Gaze_X,Gaze_Y),  # control for gaze position (Van Rij, 2019)
          data=merged_dat)
summary(m1.1)

plot_dat <- plot(m1.1)

plot_smooth(m1.1,view="Time",plot_all="condition",cond = list("Gaze_X"=960, "Gaze_Y"=540), xlab = "Time (ms)", ylab = "Pupil size (arbitrary)", main = "Partial effect of time on pupil size", ylim=c(-550, -200), legend_plot_all = "topright")

# Model checks (based on Model criticism in Van Rij, 2019)
hist(m1.1$residuals,
     xlab = "Residuals")

plot(m1.1$fitted.values,
     m1.1$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m1.1")

acf(m1.1$residuals)

# Drastic between-subject variation:
# Here we use non-linear random effects to model per subject deviations from
# the relationship between the pupil size and time (see Van Rij et al. 2019)
# for an overview about random non-linear effects see: Pedersen et al., 2019
m1.2 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=10) +
            te(Gaze_X,Gaze_Y) +  # control for gaze position
            s(Time,subject_c,bs="fs"), # m=1, so penalty on first derivative (Wood,2017) seems to diverge.
          data=merged_dat)
summary(m1.2)

plot_dat <- plot(m1.2)

plot_smooth(m1.2,view="Time",plot_all="condition",cond = list("Gaze_X"=960, "Gaze_Y"=540))
# Model checks
hist(m1.2$residuals,
     xlab = "Residuals")

plot(m1.2$fitted.values,
     m1.2$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m1.2")

acf(m1.2$residuals)

# Even simpler random effects lead to really wide CI bands:
m1.3 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=10) +
            te(Gaze_X,Gaze_Y) +  # control for gaze position
            s(subject_c,bs="re") +
            s(subject_c,condition,bs="re"),
          data=merged_dat)
summary(m1.3)

plot_dat <- plot(m1.3)

plot_smooth(m1.3,view="Time",plot_all="condition",cond = list("Gaze_X"=960,
                                                               "Gaze_Y"=540))

# Model checks
hist(m1.3$residuals,
     xlab = "Residuals")

plot(m1.3$fitted.values,
     m1.3$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m1.3")

acf(m1.3$residuals)

# How much does the smoothness assumption matter?
m1.4 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=100) +
            te(Gaze_X,Gaze_Y),  # control for gaze position
          data=merged_dat)
summary(m1.4)

plot_dat <- plot(m1.4)

plot_smooth(m1.4,view="Time",plot_all="condition",cond = list("Gaze_X"=960,
                                                              "Gaze_Y"=540))

# Model checks
hist(m1.4$residuals,
     xlab = "Residuals")

plot(m1.4$fitted.values,
     m1.4$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m1.4")

acf(m1.4$residuals)
```

### Loading and merging data from all subjects

Now for 15 seconds following a question. For "aided" only questions are
included for which the driver was in any of the two automation
modes (cruise and full).

```{r}
# Specify path and pattern
path_to_folders <- "C:/Users/Chiel/Documents/CCS/User Models/cleaned_dat/"
data_patterns <- c("_automated_question_downsampled_dat.RDS","_non_automated_question_downsampled_dat.RDS")
condition_names <- c("aided","unaided")

# collect subjects
subjects <- list.files(path_to_folders)[startsWith(list.files(path_to_folders),"UM")]

align_start <- -500
align_end <- 600000

# Check alignments
for(subj in subjects) {
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    if(min(sub_dat$Time,na.rm=T) > align_start) {
      align_start <- min(sub_dat$Time,na.rm=T)
    }
    if(max(sub_dat$Time,na.rm=T) < align_end) {
      print(max(sub_dat$Time,na.rm=T))
      print(subj)
      align_end <- max(sub_dat$Time,na.rm=T)
    }
    rm(sub_dat)
  }
  
}


# Merge data
merged_dat <- NULL
for(subj in subjects) {
  condition_index <- 1
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    cat(subj,data_pattern,"\n")

    sub_dat <- sub_dat[((sub_dat$Time >= align_start) & (sub_dat$Time <= align_end)),]
    
    if(condition_names[condition_index] == "aided") {
      # if you want to exclude on check
      # (e.g. only look at questions for aided where automation > 1 before the question)
      # uncomment this.
      # sub_dat <- sub_dat[sub_dat$val_check == 1,]
    } else {
      sub_dat$val_check <- 0
    }
    
    # Assign condition
    sub_dat$condition <- condition_names[condition_index]
    condition_index <- condition_index + 1
    sub_dat$subject_c <- subj
    
    if(length(colnames(sub_dat)) != length(colnames(merged_dat)) && subj != "UM1") {
      cat(subj, data_pattern, " had too many columns.\n")
      sub_dat <- sub_dat[,!(colnames(sub_dat) %in% c("PupilClean","PupilPrevious"))]
    }
    
    merged_dat <- rbind(merged_dat,sub_dat)
    rm(sub_dat)
  }
}

# set to factor for gams
merged_dat$condition <- as.factor(merged_dat$condition)
merged_dat$subject_c <- as.factor(merged_dat$subject_c)

# Averages plots
avgDat <- ddply(merged_dat,c("Time","condition"),summarise,
                meanPupil = mean(Pupil,na.rm=T),
                se=(sd(Pupil,na.rm = T)/length(Pupil)))


plot(0,0,type="n",ylim=c(-100,200),xlim=c(0,15000),
     ylab="Average change in pupil size",
     xlab = "Time (ms)")

colrs <- c("red","blue")
col_index <- 1
for(condition in condition_names) {
  print(condition)
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition],lwd=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] + (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] - (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  col_index <- col_index + 1
}
abline(v=0,lty=2,col="red")

```

### Fitting models

```{r}
# again build base model

m2.1 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=20) +
            te(Gaze_X,Gaze_Y),  
          data=merged_dat)
summary(m2.1)

plot(m2.1)


plot_smooth(m2.1,view="Time",plot_all="condition",cond = list("Gaze_X"=960, "Gaze_Y"=540),
            xlab = "Time (ms)", ylab = "Pupil size (arbitrary)", main = "Partial effect of time on pupil size following a question",
            ylim=c(-150, 50), legend_plot_all = "topright")


# Model checks
hist(m2.1$residuals,
     xlab = "Residuals")

plot(m2.1$fitted.values,
     m2.1$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m2.1")

acf(m2.1$residuals)

# Again drastic between subject variation:
m2.2 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=20) +
            te(Gaze_X,Gaze_Y) +  
            s(Time,subject_c,bs="fs",m=1),
          data=merged_dat)
summary(m2.2)

plot(m2.2)


plot_smooth(m2.2,view="Time",plot_all="condition",cond = list("Gaze_X"=960,
                                                            "Gaze_Y"=540))

# Model checks
hist(m2.2$residuals,
     xlab = "Residuals")

plot(m2.25$fitted.values,
     m2.25$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m2.2")

acf(m2.2$residuals)


# Again for simpler REs:
m2.3 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=20) +
            te(Gaze_X,Gaze_Y) +  
            s(subject_c,bs="re") +
            s(subject_c,condition,bs="re"),
          data=merged_dat)
summary(m2.3)

plot(m2.3)


plot_smooth(m2.3,view="Time",plot_all="condition",cond = list("Gaze_X"=960,
                                                            "Gaze_Y"=540))

# Model checks
hist(m2.3$residuals,
     xlab = "Residuals")

plot(m2.3$fitted.values,
     m2.3$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m2.3")

acf(m2.3$residuals)
```

### Loading and merging data from all subjects

Now for automated only we model the 10 seconds following an increase in automation.

```{r}
# Specify path and pattern
path_to_folders <- "C:/Users/Chiel/Documents/CCS/User Models/cleaned_dat/"
data_patterns <- c("_automated_change_downsampled_dat.RDS")
condition_names <- c("aided")

# collect subjects
subjects <- list.files(path_to_folders)[startsWith(list.files(path_to_folders),"UM")]

align_start <- -500
align_end <- 600000

# Check alignments
for(subj in subjects) {
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    if(min(sub_dat$Time,na.rm=T) > align_start) {
      align_start <- min(sub_dat$Time,na.rm=T)
    }
    if(max(sub_dat$Time,na.rm=T) < align_end) {
      print(max(sub_dat$Time,na.rm=T))
      print(subj)
      align_end <- max(sub_dat$Time,na.rm=T)
    }
    rm(sub_dat)
  }
  
}


# Merge data
merged_dat <- NULL
for(subj in subjects) {
  condition_index <- 1
  for(data_pattern in data_patterns) {
    sub_dat <- readRDS(paste0(path_to_folders,subj,"/",subj,data_pattern))
    cat(subj,data_pattern,"\n")

    sub_dat <- sub_dat[((sub_dat$Time >= align_start) & (sub_dat$Time <= align_end)),]
    
    # Assign condition
    sub_dat$condition <- condition_names[condition_index]
    condition_index <- condition_index + 1
    sub_dat$subject_c <- subj
    
    if(length(colnames(sub_dat)) != length(colnames(merged_dat)) && subj != "UM1") {
      cat(subj, data_pattern, " had too many columns.\n")
      sub_dat <- sub_dat[,!(colnames(sub_dat) %in% c("PupilClean","PupilPrevious"))]
    }
    
    merged_dat <- rbind(merged_dat,sub_dat)
    rm(sub_dat)
  }
}

# set to factor for gams
merged_dat$condition <- as.factor(merged_dat$condition)
merged_dat$subject_c <- as.factor(merged_dat$subject_c)

# Averages plots
avgDat <- ddply(merged_dat,c("Time","condition"),summarise,
                meanPupil = mean(Pupil,na.rm=T),
                se=(sd(Pupil,na.rm = T)/length(Pupil)))


plot(0,0,type="n",ylim=c(-100,50),xlim=c(0,10000),
     ylab="Average change in pupil size",
     xlab = "Time (in ms)")

colrs <- c("red")
col_index <- 1
for(condition in condition_names) {
  print(condition)
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition],lwd=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] + (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  lines(avgDat$Time[avgDat$condition == condition],avgDat$meanPupil[avgDat$condition == condition] - (1.96*avgDat$se[avgDat$condition == condition]),lty=2,col=colrs[col_index])
  col_index <- col_index + 1
}
abline(v=0,lty=2,col="red")

```

### Fitting models

```{r}
# Base Model for after change:
m3.1 <- bam(Pupil ~ s(Time) +
          te(Gaze_X,Gaze_Y),data=merged_dat)
summary(m3.1)
plot(m3.1)
plot_smooth(m3.1,view="Time",cond = list("Gaze_X"=960, "Gaze_Y"=540),
            xlab = "Time (ms)", ylab = "Pupil size (arbitrary)", main = "Partial effect of time on pupil size following an increase in automation", ylim=c(-80, 40))

# Model checks
hist(m3.1$residuals,
     xlab = "Residuals")

plot(m3.1$fitted.values,
     m3.1$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m3.1")

acf(m3.1$residuals)

# Again random non-linear REs:
m3.2 <- bam(Pupil ~ s(Time) +
         te(Gaze_X,Gaze_Y) +
         s(Time,subject_c,bs="fs",m=1),data=merged_dat)

summary(m3.2)
plot(m3.2)
plot_smooth(m3.2,view="Time",cond = list("Gaze_X"=960,
                                       "Gaze_Y"=540))

# Model checks
hist(m3.2$residuals,
     xlab = "Residuals")

plot(m3.2$fitted.values,
     m3.2$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m3.2")

acf(m3.2$residuals)

# Again simpler REs:
m3.3 <- bam(Pupil ~ s(Time) +
         te(Gaze_X,Gaze_Y) +
         s(subject_c,bs="re"),data=merged_dat)

summary(m3.3)
plot(m3.3)
plot_smooth(m3.3,view="Time",cond = list("Gaze_X"=960,
                                       "Gaze_Y"=540))

# Model checks
hist(m3.3$residuals,
     xlab = "Residuals")

plot(m3.3$fitted.values,
     m3.3$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m3.3")

acf(m3.3$residuals)
```

## References

Faraway, J. (2016). Extending the Linear Model with R: Generalized Linear, Mixed Effects and Nonparametric Regression Models, Second Edition. https://doi.org/10.1201/9781315382722

Pedersen, E., Miller, D., Simpson, G., & Ross, N. (2019). Hierarchical generalized additive models in ecology: An introduction with mgcv. PeerJ, 7, e6876. https://doi.org/10.7717/peerj.6876

van Rij, J., Hendriks, P., van Rijn, H., Baayen, R. H., & Wood, S. N. (2019). Analyzing the Time Course of Pupillometric Data. Trends in Hearing, 23, 2331216519832483. https://doi.org/10.1177/2331216519832483

Wood, S. N. (2017). Generalized Additive Models: An Introduction with R, Second Edition (2nd ed.). Chapman and Hall/CRC.
