---
title: "Different baseline analysis"
author: "Adaptive automation team"
date: "11/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(eyelinker)
library(PupilPre)
library(plyr)
library(dplyr)
library(mgcv)
library(itsadug)
source("./src/preprocessing_helper_functions.R")
```

## Purpose of this file

We received a very good comment during the presentation of our project that
our baseline calculations might obscure constant differences in the pupil size
following a multiplication problem between the two conditions. Thus we here
calculate an updated model that baselines each question with the average taken
over the entire experiment (both sessions) from the driver to which the question
can be associated.

```{r}
# Specify path to cleaned files containing folder for each subject
path_to_files <- ""
# Specify all users
users <- c("UM1","UM2","UM3","UM4","UM5","UM6","UM7","UM8")

combinedData <- NULL

# Define conditions
conditions <- c("automated","non_automated")

for(condition in conditions) {

  for(user in users){
    label <- paste0(user,"_",condition)
    
    recorded_eye <- "Left"
    
    sample_cl <- readRDS(paste0(path_to_files,user,"/",label,"_final_cleaned.RDS"))
    
    # Again check for message recording errors
    sample_cl$SAMPLE_MESSAGE <- as.character(sample_cl$SAMPLE_MESSAGE)
    sample_cl$SAMPLE_MESSAGE[startsWith(sample_cl$SAMPLE_MESSAGE,
                                        "NEXT BLOCK,QUESTION ASKED 1")] <- "QUESTION ASKED 1"
    sample_cl$SAMPLE_MESSAGE[startsWith(sample_cl$SAMPLE_MESSAGE,
                                        "ERROR MESSAGES LOST 1,QUESTION ASKED 1")] <- "QUESTION ASKED 1"
    sample_cl$SAMPLE_MESSAGE[startsWith(sample_cl$SAMPLE_MESSAGE,
                                        "NEXT BLOCK,EXPERIMENT ENDED")] <- "NEXT BLOCK"
    sample_cl$SAMPLE_MESSAGE <- as.factor(sample_cl$SAMPLE_MESSAGE)
    
    # collect all question
    questionDatCleaned <- extract_samples_before_after_msg(sample_cl,
                                                     "QUESTION ASKED",
                                                     0,
                                                     7500,
                                                     split_msg_by_block=T,
                                                     split_msg_decision=10
                                                     )
    # Fix alignment
    questionDatCleaned <- align_msg(questionDatCleaned,Msg="^QUESTION ASKED")
    questionDatCleaned$Time <- questionDatCleaned$Align
    
    # Manual baseline
    dat_automated <- readRDS(paste0(path_to_files,user,"/",user,"_automated_final_cleaned.RDS"))
    dat_non_automated <- readRDS(paste0(path_to_files,user,"/",user,"_non_automated_final_cleaned.RDS"))
    
    pupil_automated <- dat_automated$Pupil
    pupil_non_automated <- dat_non_automated$Pupil
    pupil_entire_exp <- c(as.numeric(pupil_automated),as.numeric(pupil_non_automated))
    
    # print total baseline
    print(mean(pupil_entire_exp,na.rm=T))
    questionDatCleaned$Pupil <- questionDatCleaned$Pupil - mean(pupil_entire_exp,
                                                                na.rm=T)
    
    # Downsample
    question_downsampled_data <- downsample(questionDatCleaned,
                             SamplingRate = 500,
                             NewRate = 50)
    
    if(condition == "automated") {
      question_downsampled_data$condition <- "aided"
    } else {
      question_downsampled_data$condition <- "unaided"
    }
    
    
    if(length(colnames(question_downsampled_data)) != length(colnames(combinedData))
       && user != "UM1") {
      
      cat(label, " had too many columns.\n")
      # This happens in case the wrapper function (safe_apply_manual_clean) had to intervene
      question_downsampled_data <- question_downsampled_data[,
                                                             !(colnames(question_downsampled_data)
                                                               %in% c("PupilClean","PupilPrevious"))]
    }
    
    combinedData <- rbind(combinedData,question_downsampled_data)
    
  }
  
}
combinedData$condition <- as.factor(combinedData$condition)
```

Now fit the same model fitted earlier:
```{r}
# Again - nothing significant can be concluded in any case because
# all regression assumptions are basically violated.
m2.1 <- bam(Pupil ~ condition +
            s(Time, by=condition,k=20) +
            te(Gaze_X,Gaze_Y),  
          data=combinedData)
summary(m2.1)

plot(m2.1)


plot_smooth(m2.1,view="Time",
            plot_all="condition",
            cond = list("Gaze_X"=960, "Gaze_Y"=540),
            xlab = "Time (ms)",
            ylab = "Pupil size (arbitrary)",
            main = "Partial effect of time on pupil size following a question",
            legend_plot_all = "topright")


# Model checks
hist(m2.1$residuals,
     xlab = "Residuals")

plot(m2.1$fitted.values,
     m2.1$residuals,
     xlab="Fitted",
     ylab="Residuals",
     main="m2.1")

acf(m2.1$residuals)
```